STJ2012 analysis

### 2016-07-29 analysis/STJ2012_analysis.Rmd
add TOC, organize headings

---
title: "STJ2012 analysis"
author: "Ross Cunning"
date: "July 28, 2016"
output: html_document
---

@@

## Import data

@@

## Descriptive statistics for dataset  
* Total count in OTU table: `r sum(otu_table(phy))`  
* Number of samples: `r nsamples(phy)`  
* Number of OTUs: `r ntaxa(phy)`

#### Distribution of OTU counts and reads per sample  
```{r phy_histograms, fig.height=3, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy)), plot=F)
samhist <- hist(log10(sample_sums(phy)), plot=F)

@@

#### Distribution of OTU counts and reads per sample post-filtering
```{r phy.f_histograms, fig.height=3, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
samhist <- hist(log10(sample_sums(phy.f)), plot=F)

@@

## Transform taxa count data
```{r transform}
# Convert to proportion (relative abundance)
phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
```

## Community composition in all samples
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp, fig.height=6, fig.width=10}
# Define function to plot Symbiodinium community composition
composition <- function(phy, col, legend=T) {
  samdat <- data.frame(sample_data(phy))
  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), rownames(samdat)])
  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
  shorebreaks <- which(shorebreaks==F) - 1
  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
  # Make Barplot
  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
  mtext(side=1, "Relative abundance", cex=0.7, line=1)
  # Add legend
  if (legend==T) {
    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
  }
  # Add grouping bars for Shore
  for (i in 1:length(shorebreaks)) {
    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], lwd=0.25, xpd=T)
  }
  # Add lines to separate species and species names
  for (i in 1:length(spbreaks)) {
    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
  }
}

@@

# Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
#   Calculate distances to group centroids in full-dimensional principal coordinate space
#   Implemented by vegan betadisper(), based on Anderson (2006).
# Compartments/species:
```{r beta, cache=T}
sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid")
anova(sambd)
boxplot(sambd)
sambdmeans <- aggregate(sambd$distances, by=list(group=sambd$group), FUN=mean)
sambdses <- aggregate(sambd$distances, by=list(group=sambd$group), 
                      FUN=function(x) sd(x)/sqrt(length(x)))
sampt <- permutest(sambd, pairwise = T)
sampt2 <- TukeyHSD(sambd)
library(multcompView)
saml <- multcompLetters(sampt$pairwise$permuted)#, Letters=c("e","a", "f","d","c","b"))
saml2 <- multcompLetters(sampt2$group[, 4])

>>>>

---
title: "STJ2012 analysis"
author: "Ross Cunning"
date: "July 28, 2016"
output: 
  html_document: 
    toc: yes
---

@@

# Data pre-processing
## Import dataset

@@

#### Descriptive statistics  
* Total count in OTU table: `r sum(otu_table(phy))`  
* Number of samples: `r nsamples(phy)`  
* Number of OTUs: `r ntaxa(phy)`

#### Distribution of OTU counts and reads per sample  
```{r phy_histograms, fig.height=2, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy)), plot=F)
samhist <- hist(log10(sample_sums(phy)), plot=F)

@@

#### Descriptive statistics of filtered dataset
* Total count in OTU table: `r sum(otu_table(phy.f))`  
* Number of samples: `r nsamples(phy.f)`  
* Number of OTUs: `r ntaxa(phy.f)`

#### Distribution of OTU counts and reads of filtered dataset
```{r phy.f_histograms, fig.height=2, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
samhist <- hist(log10(sample_sums(phy.f)), plot=F)

@@

## Transform count data
```{r transform}
# Convert to proportion (relative abundance)
phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
```

-----

# Data analysis
## *Symbiodinium* community composition barplots
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp, fig.height=6, fig.width=8}
# Define function to plot Symbiodinium community composition
composition <- function(phy, col, legend=T) {
  samdat <- data.frame(sample_data(phy))
  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), rownames(samdat)])
  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
  shorebreaks <- which(shorebreaks==F) - 1
  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
  # Make Barplot
  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
  mtext(side=1, "Relative abundance", cex=0.7, line=1)
  # Add legend
  if (legend==T) {
    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
  }
  # Add grouping bars for Shore
  for (i in 1:length(shorebreaks)) {
    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], lwd=0.25, xpd=T)
  }
  # Add lines to separate species and species names
  for (i in 1:length(spbreaks)) {
    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
  }
}

@@

## Beta diversity

```{r beta, cache=T}
# Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
#   Calculate distances to group centroids in full-dimensional principal coordinate space
#   Implemented by vegan betadisper(), based on Anderson (2006).
# Compartments/species:
sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid")
anova(sambd)
boxplot(sambd)
sambdmeans <- aggregate(sambd$distances, by=list(group=sambd$group), FUN=mean)
sambdses <- aggregate(sambd$distances, by=list(group=sambd$group), 
                      FUN=function(x) sd(x)/sqrt(length(x)))
sampt <- permutest(sambd, pairwise = T)
sampt2 <- TukeyHSD(sambd)
library(multcompView)
saml <- multcompLetters(sampt$pairwise$permuted)#, Letters=c("e","a", "f","d","c","b"))
saml2 <- multcompLetters(sampt2$group[, 4])
